---
title: "Fitting the EPS and EB Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting the EPS and EB Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
 
```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
 
# The MCMC fits take several minutes each, so they are not run when the
# package is built. Set this to TRUE to run the whole vignette.
run_models <- FALSE
```
 
This vignette shows how to simulate data and fit the four models in
**epsm**:
 
| | Spatial | Space-time |
|---|---|---|
| Preferential sampling | `epsm_mcmc()` | `epsm_st_mcmc()` |
| Baseline (no preferential sampling) | `ebm_mcmc()` | `ebm_st_mcmc()` |
 
Throughout, the latent Gaussian field called `S` in the code and output is
denoted $Z$ in the paper.
 
```{r load}
library(epsm)
set.seed(4649)
```
 
## Settings
 
Chain length, burn-in and thinning used below. These are for illustration;
longer runs are needed for convergence.
 
```{r settings}
NSIMS <- 1000
BURNIN <- 100
THIN <- 2
```
 
Priors are passed as a named list. `default_priors()` returns the defaults;
change individual elements as needed:
 
```{r priors}
priors <- default_priors()
# e.g. a wider prior on the preferential-sampling parameter:
# priors$PRIOR_BETA_VAR <- 4
str(priors[c("PRIOR_BETA_MEAN", "PRIOR_BETA_VAR", "PRIOR_XI_MEAN")])
```
 
The output of every model is a list of chains. `make_mcmc_list_obj()` turns
it into a standard **coda** `mcmc.list`. This helper keeps only the sampled
parameters, dropping bookkeeping variables
for these models:
 
```{r helper}
to_coda <- function(fit, drop = character(0)) {
  skip <- c("k", "pred_coords", "other_summaries", "sim_data",
            "T_n", "sigma2_T", "rho_T", drop)
  params <- setdiff(names(fit), skip)
  params <- params[!vapply(fit[params], is.null, logical(1))]
  window(make_mcmc_list_obj(list(fit), burnin = BURNIN, params = params),
         thin = THIN)
}
```
 
# Spatial models
 
## Simulated data
 
`make_sim_data_max_stations()` simulates a preferentially sampled set of
stations with one block maximum each (`obs_per_station = 1`).
 
```{r sim-spatial}
sim_dat <- make_sim_data_max_stations(beta = 2, obs_per_station = 1)
 
y <- sim_dat$y
obs_coords <- sim_dat$obs_coords
stations <- sim_dat$stations
 
# Predict at the simulated locations that were NOT sampled. Any matrix of
# coordinates can be used here.
pred_coords <- sim_dat$all_coords[(sim_dat$littlen + 1):nrow(sim_dat$all_coords), ]
 
c(stations = sim_dat$littlen, prediction_locations = nrow(pred_coords))
```
 
## 1. Extremal preferential sampling model
 
```{r epsm, eval = run_models}
mod1 <- epsm_mcmc(
  y = y,
  obs_coords = obs_coords,
  stations = stations,
  pred_coords = pred_coords,
  nsims = NSIMS,
  priors = priors
)
mod1_mcmc <- to_coda(mod1)
 
summary(mod1_mcmc[, "beta"])
plot(mod1_mcmc[, "eta.1"])
```
 
For the preferential models, predictions are made **inside** the MCMC,
because they need the full augmented field (observed plus thinned
locations). `S_pred.j` is the field at the `j`-th row of `pred_coords`:
 
```{r epsm-pred, eval = run_models}
summary(mod1_mcmc[, "S_pred.1"])
```
 
## 2. Extremal baseline model
 
Same data, without the point process for the station locations:
 
```{r ebm, eval = run_models}
modb1 <- ebm_mcmc(
  y = y,
  obs_coords = obs_coords,
  stations = stations,
  pred_coords = pred_coords,
  nsims = NSIMS,
  priors = priors
)
modb1_mcmc <- to_coda(modb1)
 
plot(modb1_mcmc[, "eta.1"])
summary(modb1_mcmc[, "S_n.1"])
```
 
For the spatial baseline model, predictions can be made inside the MCMC (as
here, via `pred_coords`) or afterwards from the posterior samples of
`S_n`, `sigma2_S` and `rho_S`.
 
# Space-time models
 
## Simulated data
 
`make_sim_data_max_years()` simulates one maximum per station per year, with
a shared field $S$ plus a year-specific field $W_t$:
 
```{r sim-st}
sim_dat <- make_sim_data_max_years(
  beta = 2,
  obs_per_station = 10, # 10 years per station
  large = FALSE,
  varying_range = FALSE
)
 
y <- sim_dat$y
obs_coords <- sim_dat$obs_coords
stations <- sim_dat$stations
years <- sim_dat$years
pred_coords <- sim_dat$all_coords[(sim_dat$littlen + 1):nrow(sim_dat$all_coords), ]
 
table(years)[1:3]
```
 
## 3. Extremal preferential sampling model, space-time
 
```{r epsm-st, eval = run_models}
mod2 <- epsm_st_mcmc(
  y = y,
  obs_coords = obs_coords,
  stations = stations,
  years = years,
  pred_coords = pred_coords,
  nsims = NSIMS,
  priors = priors
)
mod2_mcmc <- to_coda(mod2)
```
 
Year-specific effects are named `W_n_<year>.<station>`. For example, $W$ in
year 7 at the 10th station observed that year:
 
```{r epsm-st-w, eval = run_models}
plot(mod2_mcmc[, "W_n_7.10"])
```
 
**Predictions.** The shared field $S$ is predicted inside the MCMC, as in
the spatial model. The year-specific fields $W_t$ are **not**: kriging every
year at every iteration is expensive and unnecessary, so do it in
post-processing from `W_n`, `sigma2_W` and `rho_W`.
 
```{r epsm-st-pred, eval = run_models}
plot(mod2_mcmc[, "S_pred.10"])   # works
# mod2_mcmc[, "W_pred_7.10"]     # does not exist
```
 
## 4. Extremal baseline model, space-time
 
The space-time baseline has no shared field: its only spatial effect is
year-specific, and it is returned under the `W` names (`W_n`, `sigma2_W`,
`rho_W`) so it lines up with `epsm_st_mcmc()`. It does no kriging, so
`pred_coords` is ignored (with a warning) and is omitted here.
 
```{r ebm-st, eval = run_models}
modb2 <- ebm_st_mcmc(
  y = y,
  obs_coords = obs_coords,
  stations = stations,
  years = years,
  nsims = NSIMS,
  priors = priors
)
modb2_mcmc <- to_coda(modb2)
 
summary(modb2_mcmc[, c("sigma2_W", "rho_W")])
```
 
## Multiple chains
 
`make_mcmc_list_obj()` takes a list of fits, so several chains can be
combined for convergence diagnostics:
 
```{r chains, eval = run_models}
fits <- lapply(1:3, function(i) {
  ebm_mcmc(y = sim_dat$y[years == 1], obs_coords = obs_coords,
           stations = stations[years == 1], nsims = NSIMS, priors = priors)
})
chains <- make_mcmc_list_obj(fits, burnin = BURNIN, params = c("eta", "xi", "rho_S"))
coda::gelman.diag(chains)
```
